---
title: "HVP Innovation Award 2019"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: 
  - name: Yun (Renee) Zhang
    email: zhangy@jcvi.org
  - name: Aishwarya Mandava
    email: amandava@jcvi.org
  - name: Yu (Max) Qian* (*corresponding)
    email: mqian@jcvi.org
---

<br>

## The Project

[Text to be inserted.]


### Design of HVP01 study

This is a brief summary of the HVP01 study. [Text to be inserted.] For more details, please see its [clinical trial page](https://clinicaltrials.gov/ct2/show/study/NCT03083158).


### Data availability

In this project, we are going to use data from the following.

- Transcriptome profiling data
    + RNA sequencing of whole blood (bulk RNA-seq)
    + RNA sequencing of single immune cells (scRNA-seq)
- Flow cytometry data
- Hepatitis B surface antibody (anti-HBs) titer value
- Clinical metadata

For data sharing, please contact [which team?]

<br>

## The Report

This report only serves the purpose of depositing code and script for the above project. For all other purposes, please contact us directly.

### Table of contents

```{r echo=FALSE, results='asis'}
# rmd <- Sys.glob("*.Rmd")
# rmd <- rmd[rmd != "index.Rmd"]
rmd <- c("data.Rmd", "bulk.Rmd", "deconv.Rmd")
html <- sub(".Rmd", ".html", rmd)
lines <- lapply(rmd, readLines)
yaml <- lapply(lines, rmarkdown:::parse_yaml_front_matter)
cat("<ul>")
for (i in seq_along(rmd)) {
  cat(paste0("<li><a href='", html[i], "'>", yaml[[i]]$title, "</a><br/>",
             "<code>", rmd[i], "</code>", "</li>"))
}
cat("</ul>")
```

We will be using the following auxiliary packages to generate this set of reports.
```{r package, message=FALSE, warning=FALSE, catch=FALSE}
library(dplyr)
library(kableExtra)
library(ggplot2)
library(gridExtra)
library(grid)
```

<br>

## Acknowledgements

This work is funded by [Text to be inserted.] Richard Scheuermann, Brian Aevermann, UBC, ...


<br>

See our other works at [JCVI](https://www.jcvi.org/).